---
title: "prism: FIML Covariance Projection for Longitudinal Missing Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{prism: FIML Covariance Projection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The Problem

Full Information Maximum Likelihood (FIML) is the gold standard for estimating
structural equation models under missing-at-random (MAR) data. It uses all
available observations without discarding partial cases and produces consistent
parameter estimates.

However, FIML does not produce a completed dataset. The naive approach —
computing conditional expectations via `lavPredict(type = "ov")` — attenuates
the variance by 40–60%. This variance collapse destroys the statistical
properties that FIML estimation preserves.

The `prism` package bridges this gap: it produces a single completed dataset
whose covariance structure matches the FIML model-implied covariance matrix,
without variance attenuation. It supports latent growth models
(`prism_fiml()`) and arbitrary single-group structural equation models
(`prism_sem()`).

## How It Works

The pipeline has two steps:

1. **FIML Estimation:** Fit a structural equation model (growth model, CFA,
   or general SEM) via `lavaan::growth()` or `lavaan::sem()` with
   `missing = "fiml"`. Extract the model-implied covariance matrix Σ_FIML.

2. **Regularized Projection:** Starting from an initial imputation
   (user-supplied or column-mean fallback), run C++ gradient descent on the
   regularized constrained objective

   $$f(X) = \frac{1}{2N_{\text{mis}}}\|M \odot (X - X^{(0)})\|_F^2 + \frac{\lambda_\Sigma}{2p^2}\|\operatorname{Cov}(X) - \Sigma_{\text{FIML}}\|_F^2$$

   Only originally-missing cells are updated. Both terms are normalized —
   fidelity per missing cell and covariance discrepancy per matrix entry —
   so `lambda_sigma` is dimensionless and comparable across sample sizes,
   variable counts, and missingness rates. The fidelity term anchors the
   solution to the nonparametric information in the initial imputation, and
   `lambda_sigma` trades fidelity against structure. The gradient of the
   missing cells is centered at zero within each column before every step,
   so the column means of the initial imputation are preserved exactly — the
   completed data cannot drift in its first moments. Columns with a single
   missing cell are frozen by this constraint.

## Basic Usage

```{r, eval = FALSE}
library(prism)
library(lavaan)

# Define a latent growth model
model <- "
  i =~ 1*T1 + 1*T2 + 1*T3 + 1*T4
  s =~ 0*T1 + 1*T2 + 2*T3 + 3*T4
"

df <- data.frame(
  T1 = c(1.2, NA, 2.8, 3.1, 4.0),
  T2 = c(2.1, 2.5, NA, 4.0, 4.5),
  T3 = c(3.0, 3.3, 4.1, NA, 5.1),
  T4 = c(NA, 4.1, 5.0, 6.0, 6.2)
)

# Impute with covariance projection
result <- prism_fiml(df, model)
print(result)
```

## Providing an Initial Imputation

For better starting values, supply an initial imputation from any method:

```{r, eval = FALSE}
library(missRanger)
initial <- missRanger::missRanger(df, pmm.k = 5, num.trees = 100)
result <- prism_fiml(df, model, initial_imputation = initial)
```

## Tuning Parameters

- `lambda_sigma` (default 1): Structural weight of the covariance-matching
  term relative to fidelity to the initial imputation. Both losses are
  normalized (per missing cell and per covariance entry), so `lambda_sigma`
  is dimensionless and comparable across sample sizes, variable counts, and
  missingness rates; a sensitivity grid of 0.25–4 is recommended. `0`
  returns the initial imputation unchanged; larger values enforce the FIML
  structure more strongly.
- `lr` (default 1): Initial gradient descent step size; the Armijo
  backtracking line search adapts it automatically. With normalized losses
  the natural step scale is O(1).
- `tol_kkT` (default 1e-4): Stationarity tolerance on the projected-gradient
  (KKT residual) norm.
- `tol_cov` (default 1e-6): Feasibility tolerance on the covariance gap to
  the target.
- `max_iter` (default 2000): Maximum gradient descent iterations.

## Convergence Diagnostics

Every completed dataset carries a `prism_diagnostics` attribute:

```{r, eval = FALSE}
attr(result, "prism_diagnostics")
```

It reports the KKT residual `r_kkT`, the covariance feasibility gap
`feas_gap`, per-column Lagrange multiplier estimates `nu` of the mean
constraints, and a `status` classifying the termination as
`converged_feasible` (stationary and within the covariance tolerance),
`converged` (stationary with a nonzero covariance gap — the gap reflects
the fidelity/structure trade-off, not target infeasibility),
`stalled_line_search`, or `max_iter_reached`.

## Multiple Imputation

`prism_mi()` implements two-level proper multiple imputation: each completed
dataset starts from its own stochastic initial imputation (missing-value
uncertainty; by default a bootstrap-weighted `missRanger` forest) and is
projected onto a perturbed model-implied covariance drawn from the
asymptotic sampling distribution of the FIML parameters (parameter
uncertainty). It accepts any fitted `lavaan` object, including SEM fits.

## Structural Equation Models

`prism_sem()` applies the same regularized projection to any single-group
SEM. Either model syntax or a pre-fitted `lavaan` object is accepted:

```{r, eval = FALSE}
model <- "
  ind60 =~ x1 + x2 + x3
  dem60 =~ y1 + y2 + y3 + y4
  dem65 =~ y5 + y6 + y7 + y8
  dem60 ~ ind60
  dem65 ~ ind60 + dem60
"
df <- lavaan::PoliticalDemocracy
df$x1[1:10] <- NA
result <- prism_sem(df, model)
```

Ordinal, multi-group, and multilevel models are rejected with clear errors;
multi-group data can be projected group by group. Model-implied means are
not targeted: first moments are anchored to the initial imputation, exactly
as in `prism_fiml()`.
