---
title: "lagrange: FIML Covariance Projection for Longitudinal Missing Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lagrange: FIML Covariance Projection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The Problem

Full Information Maximum Likelihood (FIML) is the gold standard for estimating
structural equation models under missing-at-random (MAR) data. It uses all
available observations without discarding partial cases and produces consistent
parameter estimates.

However, FIML does not produce a completed dataset. The naive approach —
computing conditional expectations via `lavPredict(type = "ov")` — attenuates
the variance by 40–60%. This variance collapse destroys the statistical
properties that FIML estimation preserves.

The `lagrange` package bridges this gap: it produces a single completed dataset
whose covariance structure matches the FIML model-implied covariance matrix,
without variance attenuation.

## How It Works

The pipeline has two steps:

1. **FIML Estimation:** Fit a latent growth model via `lavaan::growth()` with
   `missing = "fiml"`. Extract the model-implied covariance matrix Σ_FIML.

2. **Lagrangian Projection:** Starting from an initial imputation (user-supplied
   or column-mean fallback), run C++ gradient descent to minimize:

   $$L(X) = \|\operatorname{cov}(X) - \Sigma_{\text{FIML}}\|_F^2$$

   Only originally-missing cells are updated. The gradient is computed
   analytically and masked so observed data stays frozen.

## Basic Usage

```{r, eval = FALSE}
library(lagrange)
library(lavaan)

# Define a latent growth model
model <- "
  i =~ 1*T1 + 1*T2 + 1*T3 + 1*T4
  s =~ 0*T1 + 1*T2 + 2*T3 + 3*T4
"

df <- data.frame(
  T1 = c(1.2, NA, 2.8, 3.1, 4.0),
  T2 = c(2.1, 2.5, NA, 4.0, 4.5),
  T3 = c(3.0, 3.3, 4.1, NA, 5.1),
  T4 = c(NA, 4.1, 5.0, 6.0, 6.2)
)

# Impute with covariance projection
result <- lagrange_fiml(df, model)
print(result)
```

## Providing an Initial Imputation

For better starting values, supply an initial imputation from any method:

```{r, eval = FALSE}
library(missRanger)
initial <- missRanger::missRanger(df, pmm.k = 5, num.trees = 100)
result <- lagrange_fiml(df, model, initial_imputation = initial)
```

## Tuning Parameters

- `lambda` (default 1.0): Penalty weight on the covariance constraint. Higher
  values enforce the target structure more aggressively.
- `learning_rate` (default 0.001): Gradient descent step size.
- `tol` (default 1e-6): Convergence tolerance on the Frobenius distance.
- `max_iter` (default 2000): Maximum gradient descent iterations.
